{\small} CUBEP3M: High Performance P3M N-body code

### 2012-04-17 Extendedpp.tex
More improvements

{\bf (JD, here will be one of your contribution!)}

>>>>

One of the main source of error in the calculation of the force occurs when on the smallest scales of the fine grid.
The approximation by which particles in a neighbouring mesh grid can be placed at the centre of the cell
is less accurate, which cause a maximal scatter around the exact $1/r^2$ law.
A solution to minimize this error consists in extending the pp force calculation outside a single cell,
which inevitably reintroduces a $N^2$ number of operations. Our goal is to add the flexibility to have a code
that runs slower, but produces results with a higher precision.

To allow this feature, we  have to choose how far outside a cell we want the exact pp force.  
Since the force kernels on both meshes are organized in terms of grids, the simplest way to implement this 
feature is to shut down the mesh kernels in a region of specified size, and allow the pp force to extend therein.
Concretely, these regions are constructed as cubic layers of fine mesh grids around a central cell; 
the freedom we have is to choose the number of such layers.

 To speed up the access to all particles within the domain of computation, we construct a thread safe linked list
 to be constructed and accessed in parallel by each core of the system, but this time with a head-of-chain that points to the first particle in the current fine mesh cell. We then loop over all fine grids, accessing the particles contained therein and inside each fine grid cells for which we killed the mesh kernels,
 we compute the separation and the force between each pairs and update their velocities simultaneously with Newton's third law. 
 To avoid double counting, we loop only over the fine mesh neighbours that produce non-redundant contributions. Namely, for a central cell located at 
 $(x_1, y_1, z_1)$, we only consider the neighbours $(x_2, y_2, z_2)$ that satisfy the following conditions:
 \begin{itemize}
 \item{$z_2 \ge z_1$ always}
 \item{if $z_2 = z_1$, then $y_2 \ge y_1$, otherwise we also allow $y_2 < y_1$} 
 \item{if $z_2 = z_1$ and $y_2 = y_1$, then we enforce $x_2 > x_1$}
 \end{itemize}
 The case where all three coordinates are equal is already calculated in the standard configuration of the code.

 To quantify the accuracy improvement versus computing time requirements, we performed the following test.
 We generate a set of initial conditions at a starting redshift of $z = XXX$, with a box size equal to $ XXX h^{-1}\mbox{Mpc}$,
 and with $XXX^{3}$ particles. We evolve the particles to $z=0$ with different ranges for the pp calculation, and compare 
 the resulting power spectra. For the results to be meaningful, we also need to use the same random seed for the random number generator,
 such that the only difference between different runs is the range of the pp force.
{\bf (JD, insert here your P(k) plot and a description of what you observe, in terms of accuracy improvement and real time cpu usage...)}

### 2012-04-17 Introduction.tex
More improvements

Many physical and astrophysical systems are subject to non-linear dynamics
and rely on N-body simulations to describe the evolution of bodies. 
One of the main field of application is the modelling of large scale structures, 
which are driven by the sole force of gravity. Recent observations of the cosmic microwave background
\citep{WMAP5-7} of galaxy clustering \citep{2df, SDSS, WiggleZ, BOSS} of weak gravitational lensing \citep{CFHTLS, SDSS}
and of supernovae redshift-distance relations all point towards a standard model of cosmology, in which dark energy
and collisionless dark matter occupy more than 95 per cent of the total energy density of the universe.
In such a paradigm, pure N-body code are perfectly suited to describe the dynamics, as long as we
understands where and how the baryonic fluid feeds back on the dark matter structure.
The next generation of measurements aim at constraining the cosmological parameters at the per cent level, 
and the theoretical understanding of the non-linear dynamics that govern structure formation heavily relies 
on numerical simulations.

For instance, a measurement of the baryonic acoustic oscillation (BAO) dilation scale can provide
tight constraints on the dark energy equation of scale \citep{Seo2003,2005}.
The most optimal estimates of the uncertainty requires the knowledge of the 
matter power spectrum covariance matrix, which is only accurate when measured from a large sample 
of N-body simulations \citep{RH2005, japan2,japan3}.
For the same reasons, the most accurate estimates of weak gravitational lensing signal is obtained
by propagating photons in past light cones that are extracted from simulated density fields \citep{White, Jain, Takahashi}.
{\bf (Ilian, could you say something about reionization here? Just general references as to why one needs N-body codes to perform 
accurate predictions/forecasts, etc.)}

The basic problem that is addressed with N-body codes is a time evolution of an ensemble of $N$ particles
that is subject to gravitational attraction. The brute force calculation requires $O(N^{2})$ operation, a cost that 
exceeds the memory and speed of current machines for large problems.
Solving the problem  on a mesh \citep{Hockney} reduces to $O(N\mbox{log}N)$ the number of operations,
as it is possible to solve for the particle-mesh (PM) interaction with fast Fourier transforms techniques, 
typically using high performance libraries such as {\small FFTW} \citep{FFTW}.

@@

{\small PMFAST} (\cite{PMFAST}, MPT hereafter) is one of the first code that is designed such as to optimize the PM algorithm,
both in terms of speed and memory usage. It uses a two-level mesh algorithm based on the gravity solver of \cite{TracPen2003},
The long range gravitational force is computed on a  grid four times coarser, such as to minimize the {\small MPI} communication time
and to fit in system's memory. The short range is computed locally on a finer mesh, and only the local sub-volume needs 
to be stored at a given time, allowing for {\small OPENMP}parallel computation.
This this setup enable the code to evolve very large cosmological systems both rapidly and accurately, on relatively modest clusters.
One of the main advantage of {\small PMFAST} over other PM codes is that the number of large arrays is minimized,
and the global {\small MPI} communications are cut down to the minimum: for passing particle at the beginning of each time step,
and  for computing the long range FFTs.
As described in MPT, access to particles is accelerated with the use of linked lists, deletion of `ghost' particles
in buffer zones is done at the same time as particles are passed to adjacent nodes,
and the global FFTs are performed with a slab decomposition of the volumes via a special file transfer interface, 
designed specifically to preserve a high processor load.

@@

This paper aims at presenting and quantifying these new features that make {\small CUBEP3M} one of the most competitive public N-body code.
It has already been involved in a number of scientific applications over the last few years,
spanning the field of weak lensing \citep{Sanaz_Lu, Lu_dore, Dore_Lu, Harnois_Vafaei, LuCMBLensing},  BAO
 \citep{HarnoisPen, Ngan,Zhang, Yu, HarnoisYu in prep.}, re-ionization \citep{Ilian...}, {\bf (anymore cubep3m papers to cite?)},  
and it seems relevant for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code 
The paper is structured as follow: XXXX

>>>>

Many physical and astrophysical systems are subject to non-linear dynamics
and rely on N-body simulations to describe the evolution of bodies. 
One of the main field of application is the modelling of large scale structures, 
which are driven by the sole force of gravity. Recent observations of the cosmic microwave background
\citep{WMAP5-7} of galaxy clustering \citep{2000AJ....120.1579Y, 2003astro.ph..6581C, 2009arXiv0902.4680S, 2010MNRAS.401.1429D} of weak gravitational lensing \citep{CFHTLS, SDSS}
and of supernovae redshift-distance relations all point towards a standard model of cosmology, in which dark energy
and collisionless dark matter occupy more than 95 per cent of the total energy density of the universe.
In such a paradigm, pure N-body code are perfectly suited to describe the dynamics, as long as we
understands where and how the baryonic fluid feeds back on the dark matter structure.
The next generation of measurements aim at constraining the cosmological parameters at the per cent level, 
and the theoretical understanding of the non-linear dynamics that govern structure formation heavily relies 
on numerical simulations.

For instance, a measurement of the baryonic acoustic oscillation (BAO) dilation scale can provide
tight constraints on the dark energy equation of scale \citep{Seo2003,2005}.
The most optimal estimates of the uncertainty requires the knowledge of the 
matter power spectrum covariance matrix, which is only accurate when measured from a large sample 
of N-body simulations \citep{2005MNRAS.360L..82R, 2009ApJ...700..479T, 2011ApJ...726....7T}.
For the same reasons, the most accurate estimates of weak gravitational lensing signal is obtained
by propagating photons in past light cones that are extracted from simulated density fields \citep{2003ApJ...592..699V, 2009ApJ...701..945S, 2009A&A...499...31H}.
{\bf (Ilian, could you say something about reionization here? Just general references as to why one needs N-body codes to perform 
accurate predictions/forecasts, etc.)}

The basic problem that is addressed with N-body codes is a time evolution of an ensemble of $N$ particles
that is subject to gravitational attraction. The brute force calculation requires $O(N^{2})$ operation, a cost that 
exceeds the memory and speed of current machines for large problems.
Solving the problem  on a mesh \citep{1981csup.book.....H} reduces to $O(N\mbox{log}N)$ the number of operations,
as it is possible to solve for the particle-mesh (PM) interaction with fast Fourier transforms techniques, 
typically using high performance libraries such as {\small FFTW} \citep{FFTW3}.

@@

{\small PMFAST} (\cite{2005NewA...10..393M}, MPT hereafter) is one of the first code that is designed such as to optimize the PM algorithm,
both in terms of speed and memory usage. It uses a two-level mesh algorithm based on the gravity solver of \cite{TracPen2003},
The long range gravitational force is computed on a  grid four times coarser, such as to minimize the {\small MPI} communication time
and to fit in system's memory. The short range is computed locally on a finer mesh, and only the local sub-volume needs 
to be stored at a given time, allowing for {\small OPENMP}parallel computation.
This this setup enable the code to evolve very large cosmological systems both rapidly and accurately, on relatively modest clusters.
One of the main advantage of {\small PMFAST} over other PM codes is that the number of large arrays is minimized,
and the global {\small MPI} communications are cut down to the minimum: for passing particle at the beginning of each time step,
and  for computing the long range FFTs.
As described in MPT, access to particles is accelerated with the use of linked lists, deletion of `ghost' particles
in buffer zones is done at the same time as particles are passed to adjacent nodes,
and the global FFTs are performed with a slab decomposition of the volumes via a special file transfer interface, 
designed specifically to preserve a high processor load.

@@

This paper aims at presenting and quantifying these new features that make {\small CUBEP3M} one of the most competitive public N-body code.
It has already been involved in a number of scientific applications over the last few years,
spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, Harnois_Vafaei},  BAO
 \citep{2010arXiv1008.3506Z,  2011arXiv1106.5548N, 2011arXiv1109.5746H, HarnoisYuReco}, re-ionization \citep{Ilian...}, {\bf (anymore cubep3m papers to cite?)},  
and it seems relevant for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code 
The paper is structured as follow: XXXX

### 2012-04-17 PoissonSolver.tex
More improvements

This section reviews how Poisson's equation is solved on a double-mesh configuration. 
Many parts of the algorithm are identical to {\small PMFAST}, hence we refer the reader 
to section 2 of MPT for more details. In cubep3m, the mass default assignment scheme are
a `cloud-in-cell' (cic) interpolation for the coarse grid,  and a `nearest-grid-point' interpolation 
for the fine grid \citep{Hockney}. Particle-particle interactions are not interpolated, but a sharp cutoff kills the force
for pairs closer to a tenth of a grid cell.

>>>>

This section reviews how Poisson's equation is solved on a double-mesh configuration. 
Many parts of the algorithm are identical to {\small PMFAST}, hence we refer the reader 
to section 2 of MPT for more details. In cubep3m, the mass default assignment scheme are
a `cloud-in-cell' (cic) interpolation for the coarse grid,  and a `nearest-grid-point' interpolation 
for the fine grid \citep{1981csup.book.....H}. Particle-particle interactions are not interpolated, but a sharp cutoff kills the force
for pairs closer to a tenth of a grid cell.

### 2012-04-17 Systematics.tex
Significant changes to Poisson Solver, precised the skeleton, added a few discussion shere and there...
More improvements

\section{Systematics}
Random shift, resolution, initial redshift

>>>>

\section{Systematics and Accuracy}
\label{sec:systematics}

To discuss :

1) random shift to shuffle systematic clustering at the grid scale,

2) choice of initial redshift, which can lead to large truncation error if too early, but poor inaccurate Zel'dovich if too late 
{\bf JD, here you could discuss the ra\_max tempering...}

3) limits of resolution caused by softening length,

4) Poisson shot noise if too few particles or unrelaxed systems, 5) finite box size effects (with beat coupling?)

Show plots of pairwise force (transverse and radial), density force, power spectrum, 
mention the work on covariance matrix by Ngan, Harnoisetal.

### 2012-04-17 paper.tex
More improvements

\begin{abstract}
This paper presents {\small CUBEP3M}, an upgraded version of {\small PMFAST}, 
a two-mesh gravity solver that was already among the fastest N-body codes. 
Among the principal changes, the Poisson solver includes particle-particles interactions
at the subgrid level and across neighboring cells. The volume decomposition is now cubical,
and most of the calculations on the fine mesh are {\tt openmp} parallel.
Many utilities have been added and are briefly described, including a runtime halofinder
and particle identification tags. 
We discuss the structure of the code, its accuracy, and  its scaling performance.
\end{abstract}

@@

\input{Introduction}
\input{Structure}
\input{PoissonSolver}
\input{Accuracy} %pairwise and density force, 2-points stats.
\input{ScalingPerformance}
\input{Systematics}
\input{Halos}
\input{Extensions}
\input{Conclusion}
%\input{Appendix}
\bibliographystyle{mn2e}
\bibliography{mybib3}{}
%\bibliographystyle{amsplain}

>>>>

\begin{abstract}
This paper presents {\small CUBEP3M}, an upgraded version of {\small PMFAST}, 
a two-mesh gravity solver that was already among the fastest N-body codes. 
Among the principal changes, the Poisson solver includes particle-particles interactions
at the sub-grid level and across neighbouring cells. The volume decomposition is now cubical,
and most of the calculations on the fine mesh are {\tt openmp} parallel.
Many utilities have been added and are briefly described, including a runtime halofinder
and particle identification tags. 
We discuss the structure of the code, its accuracy, and  its scaling performance.
\end{abstract}

@@

\input{Introduction}
\input{Structure}
\input{PoissonSolver}
\input{ScalingPerformance}
\input{Systematics}
\input{Halos}
\input{Extensions}
\input{Conclusion}
%\input{Appendix}
\bibliographystyle{mn2e}
\bibliography{mybib3}{}
%\bibliographystyle{amsplain}
